---
title: "Week 1 Assignment Solution"
author: "Alicia R. Chen, Yining Sun"
date: "9/1/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, collapse=TRUE, comment = NA)
knitr::opts_knit$set(root.dir = "/Users/sunyining/Dropbox (ESOC - Princeton)/Predoc Training/Course Materials/1. Week 1 Assignment/")
```

```{r, message=FALSE}
# load packages
packages <- c("tidyverse","data.table", "lubridate", 'ggplot2')
lapply(packages, library, character.only = TRUE)
```

### Data Prep

Note that tweets data has many duplicates rows. 

```{r}
### Tweets data
# Load tweets data
tweets <- fread("IRA_tweets.csv")
tweets$Date <- as.Date(tweets$Date)
tweets <- unique(tweets)
```

You need to have filtered GTD data by the relevant years and country. Also need to use the islamist_groups dataset to add an indicator for islamist. Lastly, GTD data is not balanced so need to fill in 0's across the days where no event occured.

```{r}
### GTD data
# Load and filter by Russia and 2015-2018
gtd <- fread("GTD.csv")
gtd <- filter(gtd, country_txt == "Russia")
gtd <- filter(gtd, iyear >= 2014)
gtd$Date <- as.Date(with(gtd, paste(iyear, imonth, iday,sep="-")), "%Y-%m-%d")

# Add indicator that a terrorist or islamist attack occured on these dates
gtd$terrorist <- 1
islamist_groups <- read_csv("islamist_groups.csv")
gtd$islamist <- ifelse(gtd$gname %in% c(islamist_groups$islamist_groups), 1,0)

# Balance GTD data 
full_gtd <- gtd %>%
  select(Date, terrorist, islamist) %>%
  right_join(., data.frame(Date = unique(tweets$Date)))

full_gtd[is.na(full_gtd)] <- 0

# alternative code for this
# gtd %>%
#   select(Date, terrorist, islamist) %>%
#   merge(., tweets, all.y=T)

```

Some missing values in holidays data. After investigating, they are duplicates of other rows so I chose to remove those (there are other ways to address this).

```{r}
### Russian holidays data 
holidays <- fread("Russian_Holidays.csv")
holidays <- holidays[complete.cases(holidays), ]
holidays$holiday <- 1
holidays$MonthDay <- with(holidays, paste(Month, Day, sep="-"))
```

### Constructing Panel

Note that holidays data also is unbalanced, so fill in 0's for non-holiday days once I construct the panel.

```{r}
# Merge based on date
final_panel <- tweets %>%
  plyr::join(.,full_gtd) %>%
  mutate(MonthDay = paste(month(Date,label=TRUE), format(Date, "%d"), sep="-")) %>%
  plyr::join(., select(holidays, MonthDay, Religious:holiday)) %>%
  filter(Date <= as.Date("2018-06-30") & Date >= as.Date("2015-01-01"))
final_panel[is.na(final_panel)] <- 0
```

```{r }
cat("Total observations:", nrow(final_panel))
colSums(final_panel[,c('terrorist', 'islamist', 'Religious', 'Public', 'Political', 'holiday')])
```

### Descriptive Statistics
```{r}
describe <- function(data, column){
  variable <- data[,get(column)]
  stats <- data.frame(variable = column,
                      length = length(na.omit(variable)),
                      mean = mean(na.omit(variable)),
                      median = median(na.omit(variable)),
                      min = min(na.omit(variable)),
                      max = max(na.omit(variable)),
                      sd = sd(na.omit(variable)))
  return(stats)
}

descriptives <- list()
# to get the three tweet count variables
for(i in 3:5){
  col <- colnames(final_panel)[i]
  stats <- describe(final_panel, col)
  descriptives[[i-2]] <- stats
}

df <- do.call("rbind",descriptives)
df

```

There are lots of different ways to do (b) through (d). This is just one example:

```{r}
par(mfrow=c(1,2))
hist(final_panel$tweet_count_all)
hist(log(final_panel$tweet_count_all))
```

```{r}
par(mfrow=c(1,2))
hist(final_panel$tweet_count_blm)
hist(log(final_panel$tweet_count_blm))
```

```{r}
par(mfrow=c(1,2))
hist(final_panel$tweet_count_islam)
hist(log(final_panel$tweet_count_islam))
```


```{r}
byMonth <- final_panel %>%
  group_by(month = cut(Date, "month")) %>%
    summarise(tweet_count_all = sum(tweet_count_all),
              tweet_count_islam = sum(tweet_count_islam),
              tweet_count_blm = sum(tweet_count_blm),
              holiday = sum(holiday))
long <- gather(byMonth, type, tweet_count, tweet_count_islam:tweet_count_blm)
long$month <- as.Date(long$month)

coeff = 10
ggplot(long, aes(x=month)) + 
  geom_bar(aes(y=tweet_count_all/coeff), stat = "identity",alpha=0.5,fill='lightgrey',color='lightgrey',size=.1) +
  geom_line(aes(y=tweet_count, group=type, color=type)) +
  theme_minimal() +
  scale_y_continuous(
    # Features of the first axis
    name = "Tweet Count (about BLM or Islam)",
    labels = scales::comma,
    # Add a second axis and specify its features
    sec.axis = sec_axis(~.*coeff, name="Total Tweet Count", labels = scales::comma)) +
  ggtitle("IRA Tweet Count Over Time") + 
  geom_point(aes(x = month, y = holiday, shape = factor(holiday)))

```
